\thetitle

### 2023-06-03 lang-design.tex
Fixing prec.

\begin{centered}
\begin{minipage}{\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
  [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
  [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] -> \forall t, option (P t)
(* For arbitrary [@P@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@], and we denote tmPR := tm_prect_STLC [@P@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@] *)
(* We will have following equations *)
tm_prect_STLC_unit : tmPR tm_unit      = [@$r_1$@]
tm_prect_STLC_var  : tmPR (tm_var x)   = [@$r_2$@] x
tm_prect_STLC_abs  : tmPR (tm_abs x b) = [@$r_3$@] a b (tmPR b)
tm_prect_STLC_app  : tmPR (tm_app a b) = [@$r_4$@] a (tmPR a) b (tmPR b)
\end{lstlisting}
\end{minipage}
\end{centered}

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
The reason we have this weird looking \ref{prop:detectable-partiality}
is due to a technical detail of the construction of "linv".

At one point, using "prec" to
construct "linv", we need to prove \\ \mintinline{Coq}{∀ a, (Bᵢ a → option
C) → option C}. This is generally not provable, unless we have \\
\mintinline{Coq}{option (Bᵢ a → C)} instead.\YZ{Don't you really mean
"unless we have the Detectable Partiality assumption instead"?}
\EDJreply{I am telling a more detailed technical point, i.e., how to use detectable partiality to prove linv.} To resolve that, the \ref{prop:detectable-partiality} can be used to transform the former to
the latter:
recall that, according to the semantic of inductive type,
\mintinline{Coq}{(Bᵢ a → option C)} is actually the result of using
"linv" on the substructure of "D". This function points out each of the
translation result from "D" to "C" and some of them are "None" because
of failing. For us, we only care about if \textbf{any} of the
substructure of "D" fails to translate---if any substructure fails,
then the whole translation fails and thus we should have "None" as
return.

@@

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D -> Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x -> D), (∀ (b : Bᵢ x), option (P (w b))) 
    -> option (P (dᵢ x w)))),  .., ∀ (d : D) -> option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}.

>>>>

\begin{centered}
\begin{minipage}{\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9.9}\ttfamily]
tm_prect_STLC : \forall (P : tm -> Type), [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] -> [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] ->
  [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
  [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] -> \forall t, option (P t)
(* For arbitrary [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@], and we denote [@"tmPR"@] := [@"tm_prect_STLC"@] [@$P$@] [@$r_1$@] [@$r_2$@] [@$r_3$@] [@$r_4$@] *)
(* We will have following equations *)
tm_prect_STLC_unit : tmPR tm_unit      = [@$r_1$@]
tm_prect_STLC_var  : tmPR (tm_var x)   = [@$r_2$@] x
tm_prect_STLC_abs  : tmPR (tm_abs x b) = [@$r_3$@] a b (tmPR b)
tm_prect_STLC_app  : tmPR (tm_app a b) = [@$r_4$@] a (tmPR a) b (tmPR b)
\end{lstlisting}
\end{minipage}
\end{centered}

@@

  \begin{itemize}
    \item \textlabel{(Detectable Partiality)}{prop:detectable-partiality} for each pair $x : A_i \vdash B_i(x)$,
    \begin{minted}{Coq}
      Axiom detectable:
      forall {T} {a} (f: Bᵢ a -> option T),
        {forall x, f x <> None} + {exists x, f x = None}.
    \end{minted} 
    \item Define $"C"$ as the inductive type using this list, i.e.,
    \begin{minted}{Coq}
Inductive C : Set := ... | cᵢ : forall (x : Aᵢ), (Bᵢ x -> C) -> C | ...
    \end{minted}
    \item Assume an arbitrary type \mintinline{Coq}{D : Set} with
    \begin{itemize}
      \item $n$ \textbf{functions} \mintinline{Coq}{dᵢ : forall (x : Aᵢ), (Bᵢ x -> D) -> D}  
      \item a partial recursor
      \begin{minted}[escapeinside=@@]{Coq}
prec : forall (R : Set), ..,
  (rᵢ : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)), ..,@\YZ{Can R be indexed?}\EDJreply{This part actually currently incorrect and needs an extra constraint. Let me fix other parts first}\YZreply{Is this constraint Detectable Partiality?}@
      D -> option R
      \end{minted}
      \item and $n$ (propositional) computational axioms: for all $i$, 
      \begin{minted}{Coq}
prec {R} r₁ r₂ .. rₙ (dᵢ aᵢ bᵢ) 
    = rᵢ aᵢ (fun x => prec {R} r₁ r₂ .. rₙ (bᵢ x))
      \end{minted}
    % where \mintinline{Coq}{(lift rᵢ) : (forall (x : Aᵢ), (Bᵢ x -> option R) -> option R)} is defined as expected
    \end{itemize}
  \end{itemize}
  Then there exists an \textbf{embedding} from "C" to "D".
  More concretely, we can have \mintinline{Coq}{inj : C -> D} and 
  \mintinline{Coq}{linv : D -> option C} defined using the eliminator of\/
  "C" and "prec" of\/ "D" (both acting like identity) such that
  \begin{minted}[escapeinside=@@]{Coq}
    forall c : C, linv (inj c) = some c@\YZ{Does 'linv' correpsond to 'reflect' above? If so, then the reader should be explicitly reminded of this fact rather than having to guess it.}\EDJreply{Added below. I mentioned left inverse 'linv' is the generalization}@
  \end{minted}
\end{theorem}
\begin{proof} 
First, for a fixed \mintinline{Coq}{x : Aᵢ}, given \mintinline{Coq}{fᵢ :  Bᵢ x -> T},
\\ we have \mintinline{Coq}{lift fᵢ : Bᵢ x -> option T}, simply wrapping option around the returning result.

Second, with \ref{prop:detectable-partiality} we can construct
\begin{minted}{Coq}
Definition unlift:
  forall {a} {T : Set} (f: (forall c: B a, option T)) 
                       (h : forall x, f x <> None),
    {g : (forall c: B a, T) | f = lift g}.

Theorem detect_lifted:
  forall {a} {T : B a -> Type} (f: (forall c: B a, option (T c))),
    { g | f = lift g} + {exists x, f x = None}.
\end{minted}
The idea of "unlift" is that if "f" never returns "None", then it is always a result of a "lift". "detect_lifted" is just another way of saying  \ref{prop:detectable-partiality}.

Then we can construct "inj" and "linv" as follows, using eliminator and partial recursor.
\begin{minted}{Coq}
Definition inj : C -> D :=
  C_rect (fun (_ : C) ⇒ D) 
    ...
    (fun (xᵢ : Aᵢ) (_ : Bᵢ x -> Cᵢ) (Xᵢ : Bᵢ x -> D) ⇒ dᵢ xᵢ Xᵢ)
    ... .

Definition linv : D -> option C :=
  prec C 
    ...
    (fun (aᵢ : Aᵢ) (bᵢ' : Bᵢ aᵢ → option C) ⇒
       match (detectable bᵢ') with
       | left h =>
           let (bᵢ, _) := unlift bᵢ' h in Some (cᵢ aᵢ bᵢ)
       | right _ => None
       end)
    ... .
\end{minted}
Since each ``constructor'' in "C" and "D" are well-correpsonded, when invoking "C_rect" and "prec", each case is handled similarly, and they only vary along the $\{ "A"_i \vdash "B"_i \}_{i}$ types.

The definition of "inj" and "linv" is well-formed but their behaviour is still burried in the definition. The following two equations can clarify them.

\begin{minted}{Coq}
(* inj_compute can be proved thanks to 
    the computational behaviour about C_rect built-in Coq *)
Lemma inj_compute  : forall a b,
  inj (cᵢ a b)  = dᵢ a (fun x => inj (b x)).

(* linv_compute can be proved thanks to 
    the n (propositional) computational axioms about prec *)
Lemma linv_compute : forall a b,
  linv (dᵢ a b) = 
    let b' := fun x => linv (b x) in 
    match detect_lifted b' with
    | inleft (exist _ x _) => Some (cᵢ a x)
    | inright _ => None
    end.
\end{minted}

Finally,  these two equations is enough to prove that "linv" is really a left inverse, via induction
\begin{minted}{Coq}
linv (inj (cᵢ a b)) 
= linv (dᵢ a (fun x ⇒ inj (b x)))         (* by inj_compute *)
= Some (cᵢ a (fun x ⇒ linv (inj (b x))))  (* by linv_compute *)
= Some (cᵢ a (fun x => b x))              (* by I.H. *)
= Some (cᵢ a b)                           (* by eta rule *)
\end{minted}
\end{proof}

% The reason we have this weird looking \ref{prop:detectable-partiality}
% is due to a technical detail of the construction of "linv".

% At one point, using "prec" to
% construct "linv", we need to prove \\ \mintinline{Coq}{∀ a, (Bᵢ a → option
% C) → option C}. This is generally not provable, unless we have \\
% \mintinline{Coq}{option (Bᵢ a → C)} instead.\YZ{Don't you really mean
% "unless we have the Detectable Partiality assumption instead"?}
% \EDJreply{I am telling a more detailed technical point, i.e., how to use detectable partiality to prove linv.} To resolve that, the \ref{prop:detectable-partiality} can be used to transform the former to
% the latter:
% recall that, according to the semantic of inductive type,
% \mintinline{Coq}{(Bᵢ a → option C)} is actually the result of using
% "linv" on the substructure of "D". This function points out each of the
% translation result from "D" to "C" and some of them are "None" because
% of failing. For us, we only care about if \textbf{any} of the
% substructure of "D" fails to translate---if any substructure fails,
% then the whole translation fails and thus we should have "None" as
% return.

@@

We can support partial dependent eliminator as well, but partial dependent eliminator can deduce "prec" and our "prec" is enough to contruct this left inverse for (non-indexed) inductive type. 
\begin{minted}{Coq}
  pdelim : ∀ (P : D -> Set), .., 
  (rᵢ : (∀ (x : Aᵢ) (w : Bᵢ x -> D), (∀ (b : Bᵢ x), option (P (w b))) 
    -> option (P (dᵢ x w)))),  .., ∀ (d : D) -> option (P d)
\end{minted}
To generalize to indexed inductive type, we need to base on indexed W type~\cite{martin1982constructive, morris2009indexed,jashug2017}. We have mechanized the argument for indexed inductive type and dependent eliminator, but for the case we only have one constructor in "C" and "D". This mechanization is a direct generalization of the proof given in
